---
title: "Active-Set Method for Convex QP"
output:
  html_document:
    toc: true
    toc_depth: 4 
---

# Definition of QP problem

First an object is needed that defines the QP problem with all it's constraints. Note that _I_ is just a logical vector denoting whether the constraint is an inequality (_TRUE_) or equality (_FALSE_) constraint.

```{r}
setClass('QP',
         slots = c(H='matrix', c='numeric', A='matrix', b='numeric', I='logical'), 
         validity = function(object){TRUE})

setMethod("initialize",
          'QP',
          function(.Object, H, c, A, b, I, ...){
            .Object <- callNextMethod()
            .Object@H = H
            .Object@c = c
            .Object@A = A
            .Object@b = b
            .Object@I = I
            .Object
          })
QP <- function(H, c, A = matrix(), b = vector('numeric'), I = vector()){
  QP <- new('QP', H = H, c = c, A = A, b = b, I = I)
  QP
}
```

# Active-Set Method for Convex QP

_ActiveSet_ object as defined here is a closure. It is based on Active-Set Method for Convex QP as described in Nocedal, J. e.a., _Numerical Optimization_, 2ed, 2006, p.472. Example of how to use is further down.

Note two things here.

1. Solution of step $p_k$ is not exact, so some tolerance is needed under which components of $p_k$ are considered 0. That is the _pTol_ argument. 
2. Then there is the solution of the KKT matrix. In case it is positive semi-definite, then the R base _solve()_ function can not be used. So the Moore-Penrose generalized inverse is used from _MASS_ package. _svd()_ from R base could probably also be used.

```{r}
ActiveSet <- function(QP, H = QP@H, c = QP@c, A = QP@A, b = QP@b, I = QP@I, 
                      x0, writeLog = FALSE, pTol = 1e-10){
  
  # Private members
  a <- vector('list')
  W <- vector('list')
  x <- vector('list')
  p <- vector('list')
  kCurrent <- 0
  xFinal <- NULL
  
  # Private functions
  pLog <- function(string){
    if (writeLog) print(string)
  } 
  
  activeConstraints <- function(xk){
    which(A %*% xk - b == 0)  
  }
  
  g <- function(xk){
    H %*% xk + c
  }
  
  K <- function(H, A, AtMultiplier = 1){
    top <- cbind(H, AtMultiplier * t(A))
    bottom <- cbind(A, array(0, dim = c(nrow(A), ncol(top) - ncol(A))))
    rbind(top, bottom)
  }
  
  solveKKT <- function(K, c, b){
    sol <- MASS::ginv(K) %*% c(-c, b) # Moore-Penrose generalized inverse
    structure(sol[1:length(c)], lambdas = sol[-(1:length(c))])
  }
  
  solvepk <- function(xk, Wk){
    solveKKT(K = K(H = H, A = A[Wk,,drop=FALSE]), 
             c = g(xk = xk), 
             b = rep(0, length(Wk)))
  }
  
  pk0LagrangeMultipliers <- function(xk, Wk){
    qr.solve(t(A[Wk,,drop=FALSE]), g(xk = xk)) 
  }
  
  pk0FoundX <- function(multipliers, Wk){
    # If all multipliers >= 0 for ONLY inequality constraints, 
    # then that is the solution
    all(multipliers[I[Wk]] >= 0)
  }
  
  pk0InequalityToRemove <- function(multipliers, Wk){
    # Inequality constraint that is causing trouble.
    Imultipliers <- multipliers[I[Wk]]
    IWk <- Wk[I[Wk]]
    IWk[which.min(Imultipliers)]
  }
  
  pk0Iteration <- function(k){
    L <- pk0LagrangeMultipliers(xk = x[[k]], Wk = W[[k]])
    if (pk0FoundX(multipliers = L, Wk = W[[k]])) {
      pLog(paste('x* found after', kCurrent, 'steps!'))
      xFinal <<- x[[k]]
      return(x[[k]])
    } else {
      W[[k+1]] <<- setdiff(W[[k]], pk0InequalityToRemove(L, Wk = W[[k]]))
      x[[k+1]] <<- x[[k]]
    }
  }  
  
  pkNNComputeAk <- function(xk, pk, Wk){
    Imin <- intersect(setdiff(1:length(I), Wk), which(A %*% pk < 0))
    minSet <- (b[Imin] - A[Imin,,drop=FALSE] %*% xk) /
              (A[Imin, , drop=FALSE] %*% pk)
    structure(min(1, minSet), 'blockingSet' = Imin[minSet < 1])
  }
  
  pkNNIteration <- function(k){
    a[[k]] <<- pkNNComputeAk(xk = x[[k]], pk = p[[k]], Wk = W[[k]])
    x[[k+1]] <<- x[[k]] + a[[k]] * as.vector(p[[k]])
    if (length(attr(a[[k]], 'blockingSet') > 0)) {
      W[[k+1]] <<- union(W[[k]], attr(a[[k]], 'blockingSet')[1])
    } else {
      W[[k+1]] <<- W[[k]]
    }
  }
  
  iterate <- function(k){
    p[[k]] <<- solvepk(xk = x[[k]], Wk = W[[k]])
    # Problem here is numerical precision to assume pk == 0
    if (all(abs(p[[k]]) < pTol)){ # if pk == 0
      pk0Iteration(k)
    } else { # pk != 0
      pkNNIteration(k)
    }
  }
  
  nextIteration <- function(){ 
    iterate(k = kCurrent + 1)
    kCurrent <<- kCurrent + 1
  }
  
  iterateTillFound <- function(){
    while (is.null(xFinal)){
      nextIteration()
    }
    return(xFinal)
  }
  
  # Initialize
  x[[1]] <- x0
  W[[1]] <- activeConstraints(x[[1]])
  
  # public functions
  list(nextIteration = nextIteration,
       solve = iterateTillFound,
       getWorkingSetW = function(){W},
       getStepP = function(){p},
       getStepLengthAlpha = function(){a},
       getPointsX = function(){x})
}
```

# Examples

## Example 16.2 from Nocedal (p.452-3)

```{r}
QPeEb <- QP(H = matrix(c(6,2,1,2,5,2,1,2,4), ncol = 3),
            c = -c(8,3,3),
            A = matrix(c(1,0,0,1,1,1), ncol = 3),
            b = c(3,0),
            I = c(FALSE, FALSE))
QPeEb
```

```{r}
ActiveSet(QP = QPeEb, x0 = c(3,0,0), writeLog = TRUE)$solve()
```


## Example 16.2 from Nocedal (p.475-6)

```{r}
QPeIb <- QP(H = diag(2, 2), 
            c = c(-2,-5),
            A = matrix(data = c(+1, -2,
                                -1, -2,
                                -1, +2,
                                +1, +0,
                                +0, +1), ncol = 2, byrow = TRUE),
            b = c(-2, -6, -2, 0, 0),
            I = c(TRUE, TRUE, TRUE, TRUE, TRUE))
QPeIb
```

```{r}
ActiveSet(QP = QPeIb, x0 = c(2,0), writeLog = TRUE)$solve()
```

